### 2018-01-17 submission/manuscript.Rmd
grammar corrections before a lab reading

**Results.** At 35 cycles there were significant differences between HiFi DNA polymerase for fecal samples (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001). These HiFi dependent differences in the number of OTUs could be identified as early as 20 cycles in the mock communities (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`). Chimera prevalence varied by HiFi DNA polymerase and these differences were still observed after chimera removal using VSEARCH. Further, the chimera prevalence had a strong positive correlation with the number of OTUs observed in a sample that was also not changed by chimera removal with VSEARCH.   
**Conclusions.** Due to the impact of HiFi DNA polymerase on the number of OTUs, common diversity metrics that incorporate this value could give artificially inflated numbers due to higher undetected chimeras. So, when designing 16S rRNA gene survey studies it is important to consider both the cycle number and the type of HiFi DNA polymerase that will be used since it can increase or decrease the number of OTUS that are observed.

@@

Over recent years there has been an increasing focus on standardizing methodological approaches in microbiota research [@review_Kim_2017; @review_Hugerth_2017]. In particular, a lot of the attention of this standardization push has been invested in investigating ways that 16S rRNA gene surveys can be made more replicatable [@contamination_Salter2014; @taq_Gohl_2016; @storage_Lauber_2010; @preservation_Song_2016]. Although 16S rRNA gene sequencing has been much maligned for introduced bias, many of these same considerations also affect metagenomic sequencing [@metagenomcis_bias_Nayfach_2016; @metagenomcis_bias_Costea_2017].  Between the two approaches similar bias considerations include, but are not limited to, preservation media, storage conditions, DNA extraction kit, PCR, and sequece library preparation. Thus what is identified as a source of bias in 16S rRNA gene sequencing for these overlapping considerations will also influence metagenomic sequencing results.

@@

A recent study in *Nature Biotechnology* showed that there were clear differences between normal and high fidelity (HiFi) DNA polymerase and that you could reduce error and chimera generation by optimizing the PCR protocol [@taq_Gohl_2016]. Another important component of this study was that differences, based on DNA polymerase, in the number of OTUs generated were not easily removed using the authors chosen bioinformatic pipeline [@taq_Gohl_2016]. Although it is probably not surprising that normal DNA polymerase performed worse than HiFi DNA polymerase, it is natural to extend this line of inquiry and ask whether different HiFi DNA polymerase contribute different biases to downstream sequencing results. There is some reason to think that they may since many of these HiFi DNA polymerase come from different families and may intrinsically have different error rates that cannot be completely removed with modifications [@polymerase_Ishino_2014]. In this study we critically examine if any of five different HiFi DNA polymerase introduce noticeable biases into 16S rRNA gene surveys, if this is a cycle dependent phenomenon, and whether they can be removed using a standard bioinformatic pipeline.

We amplified the 16S rRNA gene in both fecal and mock community samples using either Accuprime, Kappa, Q5, Phusion, or Platinum HiFi DNA polymerase. First, we tested if we could identify differences in the number of OTUs between the differen HiFi DNA polymerase for both the fecal and mock samples. Next, since there were differences based of HiFi DNA polymerase, we examined if a community measure such as the Bray-Curtis index would be affected. After assessing these community measures we then examined if the different HiFi DNA polymerase had different per base error rates as well as chimera prevalence. Finally, since differences in chimera prevalence, based on HiFi DNA polymerase, could not be completely removed using bioinformatic approaches we tested whether the chimera prevalence correlated well with total number of observed OTUs.

@@

***PCR Protocol:***
The five different high fidelity (HiFi) Taq DNA polymerase that were tested were AccuPrime^TM^ (ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA). The PCR cycle conditions for Platinum and Accuprime followed a previously published protocol [@protocol_Kozich_2013] (https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md). If the HiFi Taq had a specific activation time that was different then 2 minutes that was used instead. For Kappa and Q5 the protocol previously published by Gohl and colleagues was used [@taq_Gohl_2016]. For Phusion the company defined conditions were used but the same extension time as that used for Accuprime and Platinum was used.

@@

***Sequence Processing:***
The mothur software program was utilized for all sequence processing steps [@mothur_schloss_2009]. The protocol followed was similar to what has been previously publsihed [@protocol_Kozich_2013] (https://www.mothur.org/wiki/MiSeq_SOP). Two major differences from the stated protocol were the use VSEARCH instead of UCHIME for chimera detection and the use of the OptiClust algorithmn instead of average neighbor for Operational Taxonomic Unit (OTU) generation [@vsearch_Rognes_2016; @uchime_Edgar_2011; @opticlust_Westcott_2017]. Sequence error was determined using the seq.error command on mock samples after chimera removal and classification to the RDP to remove non-bacterial sequences [@mothur_schloss_2009; @vsearch_Rognes_2016; @rdp_Cole_2013].

***Statistical Analysis:***
All analysis was done with the R (v `r paste(version$major, ".", version$minor, sep = "")`) software package [@r_citation_2017]. Data tranformation and graphing was completed using the tidyverse package (v `r paste(packageVersion("tidyverse")[1])`) and colors selected using the viridis package (v `r paste(packageVersion("viridis")[1])`) [@tidyverse_2017; @viridis_citation_2017]. The total number of OTUs were analyzed using an ANOVA with a tukey post-hoc test. For the fecal samples the data was normalized to each individual by cycle number to account for the biological variation between different people. Bray-Curtis matrices were generated using mothur after 100 sub-samplings at 1000, 5000, 10000, and 15000. The distance matrix data was analyzed using PERMANOVA with the vegan package (v `r paste(packageVersion("vegan")[1])`) [@vegan_citation] and kruskal-wallis tests within R. For both error and chimera analysis, samples were tested using Kruskal-Wallis with a Dunns post-hoc test. Where applicable correction for multiple comparison utilized the Benjamini-Hochberg method [@benjamini_controlling_1995].

***Analysis Workflow:***
The total number of OTUs after sub-sampling was analyzed for both the fecal and mock community samples. Cycle dependent affects on Bray-Curits indicies were next assessed for the fecal samples looking at both overall cycle differences and within individual differences for the previous cycle (e.g. 20 versus 25, 25 versus 30, etc.). From these observations we wanted to next analyze potential reasons as to why some of these differences may have occured. First, analysis of general sequence error rate, number of sequences with an error, and base subsitution were assessed in the mock community for each Taq. After assessing these errors, the total number of chimeras was determined after sequence processing. The fecal samples were analyzed at 4 different sub-sampling levels, 1000, 5000, 10000, and 15000 while the mock community samples were analysed at 3 levels, 1000, 5000, 10000.

@@

***The Number of OTUs is Dependent on HiFi Taq Used:***
After normalization by individual, for each cycle number, we observed that for fecal samples the number of OTUs identified was dependent upon the HiFi DNA polymerase used and this difference increased as the depth of sub-sampling increased [Figure 1]. Lower cycle numbers (15-20) resulted in less differences between HiFI DNA polymerase while cycle numbers of 25, 30, and 35 had larger clearer defined differences [Figure 1]. Only 35 cycles had HiFi Taq differences that were significantly different at all sub-sampling levels (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001) [Table S1]. At sub-sampling depth of 5000 or higher 25 and 30 cycles had HiFi DNA polymerase differences (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "30x", sub_sample_level != 1000)[, "bh"]) < 0.05 & max(filter(fs_anova_summary, cycle == "25x", sub_sample_level != 1000)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S1]. Using a Tukey post-hoc test only at 35 cycles were significant differences found to be mainly driven by Platinum being different than all other Taq across sub-sampling depth (P-value `r ifelse(max(filter(fs_tukey_summary, padj < 0.05)[, "padj"]) < 0.05,  invisible("<"), invisible("check data"))` 0.05) [Table S2].

@@

***Minimal Bray-Curtis Differences are Detected by Cycle Number:***
Overall, there was very little difference between each respective 5-cycle increment (e.g. 15x vs 20x) for both fecal and mock samples and this was consistent across the different sub-samplings used [Figure 3]. Two exceptions to the low differences between 5-cycle increments can be found. The first that there seems to be large differences for fecal samples between 20x vs. 25x that is robust against sub-sampling depth [Figure 3A-B]. Second, for the mocks, where data is available, there appears to be a similar large difference detected between 20x and 25x [Figure 3D]. Regardless, by the time PCR cycles reach 25x there does not seem to be large differences in the community [Figure 3].

@@

***Sequence Error is Dependent on both Taq and Cycle Number Used:***
Differences by HiFi Taq in the median average per base error varied without a clear pattern across sub-sampling depth [Table S5]. Generally, the highest values were for the Kappa HiFI DNA polymerase [Figure 4A]. Sub-sampling depth seems to have little effect on this rate with both 5000 and 10000 sub-sampled sequences showing similar results [Figure 4B-C]. There were small differences between the various HiFi Taq at lower cycle number but larger differences at higher cycle number with Platinum having large differences between all other HiFi DNA polymerase [Figure 4B-C and Table S6].

The total seqeunces with at least one error had multiple differences at different cycle numbers and sub-sampling depth driven by large differences in Accuprime^TM^ and Platinum versus the other HiFi Taq tested [Figure S1 & Table S7 & S8]. Although Accuprime^TM^ had the lowest per base error rate it had either the largest or second largest number of sequences with at least one error regardless of cycle number or sub-sampling depth [Figure S1]. Investigation of whether there were HiFi DNA polymerase dependent effects on base subsitution found that there was no clear bias and this was indpendent of sub-sampling depth [Figure S2-S4]. Further, the variation in subsitution error seems to reduce as the sub-sampling depth increases [Figure S2-S4].

***Chimeric Sequences Correlate with OTUs and are HiFi Taq Dependent:***
After chimera removal using VSEARCH and removal of sequences that did not classify as bacteria we assessed the percentage of sequences that were still chimeric within our mock community. At all levels of sub-sampling and cycle number there were significant differences between the HiFi DNA polymerase used (P-value `r ifelse(max(filter(kruskal_chimera_prev, sub_sample_level > 500)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S9]. Using a Dunn's post-hoc test the vast majority of these differences were driven by Platinum being different then all other HiFi DNA polymerase across cycle number and sub-sampling depth [Table S10]. Generally, across sub-sampling depth and cycle number Accuprime^TM^ had the lowest chimera prevalence of all the HiFi DNA Polymerase regardless of whether pre.cluster or VSEARCH had been used [Figure 5].

For all Taqs a positive correlation was observed between chimeric sequences and number of OTUs, with this correlation being strongest for Accuprime, Platinum and Phusion HiFi DNA Polymerase [Figure 6]. In general, the R^2^ value between the number of OTUs and chimeric sequences did not change from the use of pre.cluster and VSEARCH [Figure 6]. Taken together this data suggests that a strong correlation exists between the number of OTUs and the prevalence of chimeric sequences. Kappa had the highest per base error rate and the lowest correlations between the number of OTUs and chimeric sequences across sub-sampling depth [Figure 4 & 6].

@@

Our observations build upon previous studies [@taq_Gohl_2016] by showing that even different HiFi DNA Polymerase have significant differences in the number of OTUs and that the changes to total OTUs correlate with chimeras not removed after sequence processing [Figure 1-2 & 5]. This is important since many diversity metrics rely on the total number of OTUs as part of their calculations and changes to the total number of OTUs could drastically change the results as well as the findings. Although the attention has mostly been on standardizing and improving collection and extraction methods [@contamination_Salter2014] our observations show that independent of this consideration HiFi DNA polymerase can have a noticeable affect on the OTUs generated that can be found across sub-sampling depth and PCR cycle number [Figure 2-4]. These differences were observed in high biomass samples, where biases introduced by such components like kit contamination have less of an effect, suggesting that these differences may be exacerbated in low biomass samples.

Although we did not observe strong differences, based on cycle number, using the Bray-Curtis index the data suggests that there may be differences between 15 cycles and higher cycle numbers, such as 30x, that are commonly used. What is most interesting is that there was no difference within individuals between corresponding 5 cycle increments (e.g. 15 to 20, 20 to 25, etc.). Conversely, this may be due to low power and on observation there does seem to be a trend that 20x and 25x communities are very different [Figure 3]. This finding, in conjunction with the PERMANOVA results, suggest that cycle number can change bacterial community calculations but that these differences are minimal once at least 25 cylces are reached. Increasing the sub-sampling depth, for some DNA polymerase may reduce some of these observed community differences at lower cycle numbers.

Increasing the cycle number also exacerbated chimera prevalence differences between the different HiFi DNA polymerases [Figure 5]. The chimera prevalence was very strongly correlated with the number of OTUs and this value is relied upon heavily for many different downstream community metric calculations. However, Bray-Curtis analysis with PERMANOVA showed few differences based on DNA polymerase. Since it is possible that many of the increased number of OTUs, generated as cycle number increases, are not highly abundant that the Bray-Curtis index is able to successfully downweight these respective OTUs [@bc_index_Minchin1987]. So, choice of downstream diversity metric could be an important consideration to mitigate these observed changes due to high chimera prevalence in HiFi DNA polymerase such as Platinum.

>>>>

**Results.** At 35 cycles there were significant differences between HiFi DNA polymerase for fecal samples (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001). These HiFi dependent differences in the number of OTUs could be identified as early as 20 cycles in the mock communities (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`). Chimera prevalence varied by HiFi DNA polymerase and these differences were still observed after chimera removal using VSEARCH. Additionally, the chimera prevalence had a strong positive correlation with the number of OTUs observed in a sample and was also not changed by chimera removal with VSEARCH.

**Conclusions.** Due to the impact of HiFi DNA polymerase on the number of OTUs, common diversity metrics that incorporate this value could give artificially inflated numbers due to higher undetected chimeras. So, when designing 16S rRNA gene survey studies it is important to consider both the cycle number and the type of HiFi DNA polymerase that will be used since it can increase or decrease the number of OTUS that are observed.

@@

Over recent years there has been an increasing focus on standardizing methodological approaches in microbiota research [@review_Kim_2017; @review_Hugerth_2017]. In particular, a lot of the attention of this standardization push has been invested in investigating ways that 16S rRNA gene surveys can be made more replicatable [@contamination_Salter2014; @taq_Gohl_2016; @storage_Lauber_2010; @preservation_Song_2016]. Although 16S rRNA gene sequencing has been much maligned for introduced bias, many of these same considerations also affect metagenomic sequencing [@metagenomcis_bias_Nayfach_2016; @metagenomcis_bias_Costea_2017].  Between the two approaches similar bias considerations include, but are not limited to, preservation media, storage conditions, DNA extraction kit, PCR, and sequence library preparation. Thus what is identified as a source of bias in 16S rRNA gene sequencing for these overlapping considerations will also influence metagenomic sequencing results.

@@

A recent study in *Nature Biotechnology* showed that there were clear differences between normal and high fidelity (HiFi) DNA polymerase and that you could reduce error and chimera generation by optimizing the PCR protocol [@taq_Gohl_2016]. Another important component of this study was that differences, based on DNA polymerase, in the number of OTUs generated were not easily removed using the authors chosen bioinformatic pipeline [@taq_Gohl_2016]. Although it is probably not surprising that normal DNA polymerase performed worse than HiFi DNA polymerase, it is natural to extend this line of inquiry and ask whether different HiFi DNA polymerase contribute different biases to downstream sequencing results. There is some reason to think that this may be the case since many of these HiFi DNA polymerase come from different families and may intrinsically have different error rates that cannot be completely removed with modifications [@polymerase_Ishino_2014]. In this study we critically examine if any of five different HiFi DNA polymerase introduce noticeable biases into 16S rRNA gene surveys, if this is a cycle dependent phenomenon, and whether they can be removed using a standard bioinformatic pipeline.

We amplified the 16S rRNA gene in both fecal and mock community samples using either Accuprime, Kappa, Q5, Phusion, or Platinum HiFi DNA polymerase. First, we tested if we could identify differences in the number of OTUs between the differen HiFi DNA polymerase for both the fecal and mock samples. Next, since there were differences based on HiFi DNA polymerase, we examined if a community measure such as the Bray-Curtis index would be affected. After assessing these community measures we then examined if the different HiFi DNA polymerase had different per base error rates as well as chimera prevalence. Finally, since differences in chimera prevalence, based on HiFi DNA polymerase, could not be completely removed using bioinformatic approaches we tested whether the chimera prevalence correlated well with total number of observed OTUs.

@@

***PCR Protocol:***
The five different high fidelity (HiFi) Taq DNA polymerase that were tested were AccuPrime^TM^ (ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA). The PCR cycle conditions for Platinum and Accuprime followed a previously published protocol [@protocol_Kozich_2013] (https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md). If the HiFi Taq had a specific activation time that was different then 2 minutes, the specified activation was used instead. For Kappa and Q5 the protocol previously published by Gohl and colleagues was used [@taq_Gohl_2016]. For Phusion the company defined conditions were used but the same extension time as that used for Accuprime and Platinum was used.

@@

***Sequence Processing:***
The mothur software program was utilized for all sequence processing steps [@mothur_schloss_2009]. The protocol followed was similar to what has been previously publsihed [@protocol_Kozich_2013] (https://www.mothur.org/wiki/MiSeq_SOP). Two major differences from the stated protocol were the use of VSEARCH instead of UCHIME for chimera detection and the use of the OptiClust algorithmn instead of average neighbor for Operational Taxonomic Unit (OTU) generation at 97% similarity [@vsearch_Rognes_2016; @uchime_Edgar_2011; @opticlust_Westcott_2017]. Sequence error was determined using the seq.error command on mock samples after chimera removal and classification to the RDP to remove non-bacterial sequences [@mothur_schloss_2009; @vsearch_Rognes_2016; @rdp_Cole_2013].

***Statistical Analysis:***
All analysis was done with the R (v `r paste(version$major, ".", version$minor, sep = "")`) software package [@r_citation_2017]. Data tranformation and graphing was completed using the tidyverse package (v `r paste(packageVersion("tidyverse")[1])`) and colors selected using the viridis package (v `r paste(packageVersion("viridis")[1])`) [@tidyverse_2017; @viridis_citation_2017]. Differences in the total number of OTUs were analyzed using an ANOVA with a tukey post-hoc test. For the fecal samples the data was normalized to each individual by cycle number, to account for the biological variation between different people. Bray-Curtis matrices were generated using mothur after 100 sub-samplings at 1000, 5000, 10000, and 15000. The distance matrix data was analyzed using PERMANOVA with the vegan package (v `r paste(packageVersion("vegan")[1])`) [@vegan_citation] and kruskal-wallis tests within R. For both error and chimera analysis, samples were tested using Kruskal-Wallis with a Dunns post-hoc test. Where applicable correction for multiple comparison utilized the Benjamini-Hochberg method [@benjamini_controlling_1995].

***Analysis Workflow:***
The total number of OTUs after sub-sampling was analyzed for both the fecal and mock community samples. Cycle dependent affects on Bray-Curits indicies were next assessed for the fecal samples looking at both overall cycle differences and within individual differences for the previous cycle (e.g. 20 versus 25, 25 versus 30, etc.). Based on these observations we wanted to next analyze potential reasons as to why some of these differences may have occured. First, analysis of general sequence error rate, number of sequences with an error, and base subsitution were assessed in the mock community for each DNA polymerase. After assessing these errors, the total number of chimeras was determined after sequence processing. For the community based measures, the fecal samples were analyzed at 4 different sub-sampling levels, 1000, 5000, 10000, and 15000 while the mock community samples were analysed at 3 levels, 1000, 5000, 10000.

@@

***The Number of OTUs is Dependent on HiFi DNA Polymerase Used:***
After normalization by individual, for each cycle number, we observed that for fecal samples the number of OTUs identified was dependent upon the HiFi DNA polymerase used and this difference increased as the depth of sub-sampling increased [Figure 1]. Lower cycle numbers (15-20) resulted in less differences between HiFI DNA polymerase while cycle numbers of 25, 30, and 35 had larger clearer defined differences [Figure 1]. Only 35 cycles had HiFi DNA polymerase differences that were significantly different at all sub-sampling levels (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001) [Table S1]. At sub-sampling depth of 5000 or higher 25 and 30 cycles had HiFi DNA polymerase differences (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "30x", sub_sample_level != 1000)[, "bh"]) < 0.05 & max(filter(fs_anova_summary, cycle == "25x", sub_sample_level != 1000)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S1]. Using a Tukey post-hoc test only at 35 cycles were significant differences found to be mainly driven by Platinum being different than all other DNA polymerases across sub-sampling depth (P-value `r ifelse(max(filter(fs_tukey_summary, padj < 0.05)[, "padj"]) < 0.05,  invisible("<"), invisible("check data"))` 0.05) [Table S2].

@@

***Minimal Bray-Curtis Differences are Detected by Cycle Number:***
Overall, there was very little difference between each respective 5-cycle increment (e.g. 15x vs 20x) for both fecal and mock samples and this was consistent across the different sub-samplings used [Figure 3]. Two exceptions to the low differences between 5-cycle increments can be found. The first, that there seems to be large differences for fecal samples between 20x vs. 25x that is robust against sub-sampling depth [Figure 3A-B]. Second, for the mocks, where data is available, there appears to be a similar large difference detected between 20x and 25x [Figure 3D]. Regardless, by the time PCR cycles reach 25x there does not seem to be large differences in the community between the same sample [Figure 3].

@@

***Sequence Error is Dependent on both HiFi DNA Polymerase and Cycle Number Used:***
Differences by HiFi DNA polymerase in the median average per base error varied without a clear pattern across sub-sampling depth [Table S5]. Generally, the highest values were for the Kappa HiFI DNA polymerase and the pre-cluster step and chimera removal by VSEARCH had little effect on this rate [Figure 4]. Generally, there were small differences between the various HiFi DNA polymerase at lower cycle numbers and larger differences at higher cycle number with Platinum having large differences between all other HiFi DNA polymerase [Figure 4B-C and Table S6].

The total seqeunces with at least one error had multiple differences at different cycle numbers and was mostly alleviated by the use of the pre.cluster step [Figure S1]. Major differences before this pre.cluster step were driven by large differences in Accuprime^TM^ and Platinum versus the other HiFi DNA polymerase tested [Figure S1 & Table S7 & S8]. Although Accuprime^TM^ had the lowest per base error rate it had the largest number of sequences with at least one error, regardless of cycle number or sub-sampling depth [Figure S1]. However, this increased number of sequences with an error can be drastically lowered with existing bioinformatic approaches [Figure S1]. Investigation of whether there were HiFi DNA polymerase dependent effects on base subsitution found that there were generally no biases in the types of subsitution made [Figure S2].

***Chimeric Sequences Correlate with OTUs and are HiFi DNA Polymerase Dependent:***
After chimera removal using VSEARCH and removal of sequences that did not classify as bacteria we assessed the percentage of sequences that were still chimeric within our mock community. At all levels of sub-sampling and cycle number there were significant differences between the HiFi DNA polymerase used (P-value `r ifelse(max(filter(kruskal_chimera_prev, sub_sample_level > 500)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S9]. Using a Dunn's post-hoc test the vast majority of these differences were driven by Platinum being different then all other HiFi DNA polymerase across cycle number and sub-sampling depth [Table S10]. Generally, across sub-sampling depth and cycle number Accuprime^TM^ had the lowest chimera prevalence of all the HiFi DNA Polymerase regardless of whether pre.cluster or VSEARCH had been used [Figure 5].

For all DNA polymerase, a positive correlation was observed between chimeric sequences and number of OTUs, with this correlation being strongest for Accuprime, Platinum and Phusion HiFi DNA Polymerase [Figure 6]. In general, the R^2^ value between the number of OTUs and chimeric sequences did not change from the use of pre.cluster and VSEARCH [Figure 6]. Taken together this data suggests that a strong correlation exists between the number of OTUs and the prevalence of chimeric sequences. Additionally, Kappa had the highest per base error rate and the lowest correlations between the number of OTUs and chimeric sequences across sub-sampling depth [Figure 4 & 6].

@@

Our observations build upon previous studies [@taq_Gohl_2016] by showing that even different HiFi DNA polymerase have significant differences in the number of OTUs and that the changes to total OTUs correlate with chimeras not removed after sequence processing [Figure 1-2 & 5]. This is important since many diversity metrics rely on the total number of OTUs as part of their calculations and changes to the total number of OTUs could drastically change the results, as well as the findings. Although the attention has mostly been focused on standardizing and improving collection and extraction methods [@contamination_Salter2014] our observations show that independent of this consideration HiFi DNA polymerase can have a noticeable affect on the OTUs generated that can be found across sub-sampling depth and PCR cycle number [Figure 2-4]. These differences were observed in high biomass samples, where biases introduced by such components like kit contamination may have less of an effect, suggesting that these differences may be exacerbated in low biomass samples.

Although we did not observe strong differences, based on cycle number, using the Bray-Curtis index the data suggests that there may be differences between 15 cycles and higher cycle numbers, such as 30x, that are commonly used. Additionally, there was no difference within individuals between corresponding 5 cycle increments (e.g. 15 to 20, 20 to 25, etc.). Conversely, this may be due to low power and on observation there does seem to be a trend that 20x and 25x communities are very different from each other [Figure 3]. This finding, in conjunction with the PERMANOVA results, suggest that cycle number can change bacterial community calculations but that these differences are minimal once 25 cylces are reached. Increasing the sub-sampling depth, for some DNA polymerase, may reduce some of these observed community differences at lower cycle numbers.

Increasing the cycle number also exacerbated chimera prevalence differences between the different HiFi DNA polymerases [Figure 5]. The chimera prevalence was strongly correlated with the number of OTUs and this value is relied upon heavily for many different downstream community metric calculations. However, Bray-Curtis analysis with PERMANOVA showed few differences based on DNA polymerase. Since it is possible that many of the increased number of OTUs, generated as cycle number increases, are not highly abundant allowing the Bray-Curtis index to be able to successfully downweight these respective OTUs [@bc_index_Minchin1987]. So, choice of downstream diversity metric could be an important consideration in helping to mitigate these observed changes due to high chimera prevalence in HiFi DNA polymerase.
